Contracts for unions and intersections are hard for Nickel

### 2021-05-26 paper.tex
Rework the intro and the contract part

Build systems, package managers, operating systems, cloud infrastructure,
continuous integration and web services\resolved{Also continuous integration}
are examples of modern complex software systems which require an extensive and
non-trivial configuration. To handle this complexity, the declarative approach
has become more and more prominent: instead of applying a sequence of
transformations to a system in order to reach a desired state, this state is
described as static textual data. Some tool is then in charge of applying the
necessary transformations. This approach makes the configuration persistent,
reproducible, analyzable and simulable. Text-based configurations are led to
play an important role in critical aspects of software engineering, including
security, availability, and maintainability.

However, static text-based configurations show their limits beyond a given size
and complexity. Data cannot be transformed, combined, shared, nor validated.
This lack of expressivity results at best in boilerplate and duplication of
information, but it can also make data invalid or inconsistent. Correctly
configuring a modern system is hard and failures may have substantial negative
consequences.

This is why the authors have been developing the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight and easy to write while providing support for writing correct code
and generating valid data. Nickel originated as a rethink of the language of the
Nix package manager, used to describe one of the biggest sets of packaged
software in existence\cite{repology}.  Nickel targets build systems and cloud deployment
tools (Kubernetes, Terraform or NixOps) as well.

@@

The end goal being data generation, Nickel is a dynamic, pure and functional
language. Our goal is that writing a simple configuration in Nickel feels as
easy as writing JSON or YAML. Unlike JSON, though, Nickel anticipates large
configurations by being programmable, and by providing means of ensuring
correctness.

In this context, correctness means two things:
\begin{enumerate}
    \item The program runs without errors. For example, an error is raised when
        evaluating the nonsensical division \nickel{1 / "ftp"}.
    \item The program generates valid data, i.e. data that conforms to a schema
        (imposed by the system consuming the configuration). For example, when
        configuring a systemd service, the schema may require the presence of a
        field \lstinline+port+ that holds a valid port number.
\end{enumerate}

A traditional answer is to use types, and in particular static types for (1).
However, Nickel programs are of a special kind: they are usually simple,
terminating programs that run on fixed inputs. Errors of type (1) will show up
at evaluation anyway, or will not matter (dead code). For validation (2),
traditional types are not adapated: statically checking common requirements
like being a valid port number or a valid URL is, while not technically
impossible, really hard to do.

In a dynamic language, it is natural and easier to validate the data at runtime,
\textit{a posteriori}. This is the principle of dynamic typing. But dynamically
typed languages usually only check data when used in a primitive operation, like
addition. In Nickel, the interpreter could not have a clue that a field
\nickel{port}

In the configuration setting, there is a
singular dilemma with respect to typing: since a program is run on fixed inputs
and is expected to terminate, any relevant type error will show up at
evaluation.  Why bother with the complexity of a static type system? On the
other hand, as the configurated systems become more and more complex, so do
configuration codebases. Medium-sized to big codebases necessarily , static
types become attractive again.  For reusable code — that is, library functions
—, static types are specifically adapted, and bring all the usual benefits of
early error detection, code robustness, better code structure, documentation,
and so on.

Enforcing pre- and post-conditions at runtime is a widely established practice.
In their foundational paper\cite{FindlerFelleisenHOContracts}, Findler and
Felleisen introduce a principled approach to run-time assertion checking that
nicely supports higher-order functions and introduces the notion of blame, which
is crucial to good error reporting. It became apparent later that their
contracts are closely related to the type casts introduced by gradual typing,
modulo blame\unsure{What does ``modulo blame'' means here?}: both \cite{FindlerMultiLang} and \cite{FelleisenInterLang} see the
value of contracts as a safe interface between typed and untyped code. In
\cite{WellTypedBlamed}, the authors precisely introduce a system integrating
gradual typing with contracts \textit{à la Findler \& Felleisen}.\unsure{The
historical bit should probably be moved to related works eventually} Nickel
adopts a similar type system, with both statically typed terms, dynamically
typed terms, and first-class contracts. In "first-class contracts", we include
the ability to write user-defined contracts. Because built-in types are too
simplistic for the purpose of data validation, the ability to write a small
validation function for, say, a port number, to have it immediately available as
a contract that can be composed with other type constructors\resolved{Needs code
to illustrate.}, is in our opinion one of the main appeal of Nickel's contracts
system. An example is given in Figure~\ref{fig:contract-composition}. The first
line defines a port contract from a predicate and use it in the definition of a
larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
contract checks that a value is a record with the following shape: it must have
a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
a \emph{schema}, used to validate the final data. The \lstinline+|+ operator
entails that a value respect a contract, akin to \lstinline+:+ for types. We
assume that a \lstinline+Host+ contract has been defined in a similar
way.\unsure{What's missing in this section is an example of how
  contracts differ from standard dynamic type checking on some
  function. Namely, the quality of error messages.}

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nickel-lang.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\medskip
\small
Defines a port contract from a predicate and use it in the definition of a
\lstinline+Configuration+ record contract. \lstinline+Configuration+ plays the
role of a \emph{schema} for the final data. The \lstinline+|+ operator applies a
contract to a value, similar to `:` for static types. We assume that a
\lstinline+Host+ contract has also been defined.
\end{figure}

>>>>

Build systems, package managers, operating systems, cloud infrastructure,
continuous integration and web services\resolved{Also continuous integration}
are examples of modern complex software systems which require an extensive and
non-trivial configuration. To handle this complexity, the declarative approach
has become more and more prominent: instead of applying a sequence of
transformations to a system in order to reach a desired state, this state is
described as static textual data. Some tool is then in charge of applying the
necessary transformations. This approach makes the configuration persistent,
reproducible, versionable and subject to static analysis. Text-based
configurations are thus led to play an important role in critical aspects of
software engineering, including security, availability, and maintainability.

However, static text-based configurations show their limits beyond a certain
size and complexity. Data cannot be transformed, combined, shared, nor validated
(without resorting to separate schemas and tools).  This lack of expressivity
results at best in boilerplate and duplication of information, but it can also
make data invalid or inconsistent.  Correctly configuring a modern system is ard
and failures may have substantial negative consequences.

This is why the authors have been developing the Nickel
language\cite{NickelRepo}. Nickel is a configuration programming language,
meaning that a Nickel program evaluates to a value that must be
straightforwardly serializable to JSON, YAML or alike. Nickel aims at being
lightweight and easy to write while providing support for writing correct code
and generating valid data. Nickel originated as a rethink of the language of the
Nix package manager, used to describe one of the biggest sets of packaged
software in existence\cite{repology}.  Nickel targets build systems, cloud
deployment tools (Kubernetes, Terraform or NixOps), or anything requiring a
non-trivial configuration as well.

@@

The end goal being data generation, Nickel is a dynamic, pure and functional
language. Our goal is that writing a simple configuration in Nickel feels as
easy as writing JSON or YAML. Unlike JSON, though, Nickel anticipates large
configurations by being programmable, and by providing means of ensuring
both correctness and validity:

\begin{description}
    \item[(correctness)] The program runs without errors. Errors may be type
        errors, raised for example when evaluating the nonsensical division
        \nickel{1 / "ftp"}. More generally, they can any the violation of a pre-
        or post-condition. When such an error cannot be prevented, we want at
        least early failures and relevant error messages.
    \item[(validity)] The program generates valid data, i.e. data that
        conforms to a schema (imposed by the system consuming the
        configuration). For example, when configuring a systemd service, the
        schema may require the presence of a field \lstinline+port+ that holds a
        valid port number.
\end{description}

A traditional answer is to use static types, in particular static types for
correctness. However, Nickel programs are of a special kind: they are usually
simple, terminating programs that run on fixed inputs. Correctness errors will
show up at evaluation anyway, or will not matter (dead code). With respect to
validation, traditional types seem too limited. Validation commonly involves
checking requirements such as being a valid port number or a valid URL which
are, while not technically impossible, really hard to check statically.

In a dynamic language, it is natural and easier to validate data at runtime.
This underlies the approach of dynamic typing. But dynamically typed languages
usually only check the same simple predicates as static types (being a number,
being a string, etc.), and solely for data that are arguments to a primitive
operation. For validation, we want to check arbitrary predicates on arbitrary
values. This is provided in a basic form by assertions, available in most
languages out there. An assertion is an expression that verifies that a
predicate holds, or abort the program otherwise.  While assertions are fine for
simple sanity checks, they can quickly becomes verbose, clumsy to compose and to
reuse, and provide very limited context attached to an error. It turns out there
is a better tool than plain assertions: contracts.

\subsection*{Contracts}

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely supports
higher-order functions. If the user has written an \nickel{Even} contract for
number, they can use it right away to form a \nickel{#Even -> #Int} contract, as
illustrated in Figure~\ref{fig:contract-higher-order}. This contracts will check
at each call that the argument is an even number (pre-condition) and that the
return value is a integer (post-condition). The \nickel{#} character
distinguishes contracts in Nickel. Contracts also introduce the notion of blame
which is crucial to good error reporting in presence of higher-order contracts
and other composed contracts.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Even = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 2 == 0) in
let Int = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 1 == 0) in
{divBy2 | #Even -> #Int = fun x => x/2}
\end{lstlisting}
  \end{center}
\caption{Higher-order contract}
\label{fig:contract-higher-order}
\end{figure}

In our opinion, one of the main appeal of Nickel is this ability to write a
simple validation function for, say, a port number, and have it immediately
available as a contract that can be used to construct a other contracts by
writing natural type-like expressions. Another example is given in
Figure~\ref{fig:contract-composition}. The first line defines a port contract
from a predicate and use it in the definition of a larger
\lstinline+Configuration+ record contract. The \lstinline+Configuration+
contract checks that a value is a record with the following shape: it must have
a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
a \emph{schema} used to validate the final data, which is exactly what we were
looking for in the previous section. Not only it plays the role of a schema, but
it also \emph{looks} like one, as opposed to a less readable plain validation
function. The \lstinline+|+ operator entails that a value respect a contract,
akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
been defined in a similar way. \unsure{What's missing in this section is an
example of how contracts differ from standard dynamic type checking on some
function. Namely, the quality of error messages.}

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | #Host,
  port | #Port,
} in
{host = "nickel-lang.org", port = 80}
| #Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}

Contracts have another perk: they correspond to the type casts of gradual
typing, a discipline for mixing static and dynamic typing in the same language.
As such, they are a precious ingredient for handling the interaction of
statically typed and dynamically typed code. While we focused on the writing of
configuration, Nickel is a proper programming language that is also designed to
write generic and reusable library code. There, the typing trade-off is
different, and static typing is more adapted. Nickel has a gradual type system,
which is yet another argument for having contracts. Let us end this section by
stressing the following points:
\begin{itemize}
    \item \textbf{Even in a purely dynamically typed language, contracts are an
        appealing feature for validation and correctness}. The gradual
        typing part deserved to be mentioned, but is is out of the scope of this
        paper. 
    \item \textbf{User-defined contracts are essential for validation}. Some contract
        systems (as the one of~\ref{RootCauseOfBlame} that will be extensively
        discussed later in the paper) do not feature user-defined contracts.
        Rather, a fixed set of built-in contracts such as \nickel{Number} or
        \nickel{String} are provided, very much like base types. The absence of
        user-defined contracts is understandable when using contracts as a
        dynamic typing discipline for correctness. But a system without
        user-defined contracts is limited in the same way as static typing with
        respect to validation.
\end{itemize}
